Think deep for reef resilience

### 2025-08-13 paper/paper_draft.qmd
Change mathematical notation and update clust assignment/regression info

Using median coral cover timeseries for each reef we calculated the number of years each
reef maintains coral cover above a given percent threshold. The threshold represents the
level of live coral cover necessary to maintain a positive carbonate budget (denoted here as
$\beta_{0}$ to indicate the budget equilibrium). Greater number of years above $\beta_{0}$
therefore suggests greater resilience and opportunity for recovery.

Recent analyses suggest a general threshold of \~17% live coral cover in the case of the GBR
[@desbiens2021]. Given in situ carbonate budgets are influenced by (1) a range of processes;
(2) the variety of approaches that could be used to measure those processes; and (3) the
uncertainty surrounding these, it would be unreasonable to assume all reefs adhere to a
single precise $\beta_{0}$ value. Concurrently, it is difficult to ascertain a suitable
$\beta_{0}$ value for each individual reef.

Given the uncertainties present, the proportion of modelled reefs that display capacity to
maintain coral cover above a given $\beta_{0}$ threshold is examined. An indicative range of
10 to 20% is used (inclusive, with a step size of 1%). The range was informed by Perry et
al., [-@perry2015] and Desbians et al., [-@desbiens2021] providing plausible indications of
the lower and upper bounds respectively. Robustness of the conclusions drawn to the
uncertainty in the precise value of $\beta_{0}$ for all reefs can then be assessed by using
the indicative range.

The relationship between the number of years above $\beta_{0}$ and median reef depth was
then assessed with Spearman's rank correlation. Spearman's correlation coefficient was
adopted as reef depth and connectivity strength distributions are expected to be non-normal
and highly skewed. Furthermore, reefs are naturally of different types and experience
different conditions, and so the monotonic trends (whether the relationship consistently
increase or decrease) rather than the exact quantative relationship was of interest for this
analysis.

@@

To identify groups of reefs that are resilient, intermediate, and non-resilient, we
clustered reefs into three clusters based on their temporal behaviour as indicated by coral
cover over time. The Complexity-Invariant Distance (CID) between each timeseries was
calculated to identify timeseries that behave similarly [@batista2014]. The CID corrects the
Euclidean distance between pairs of timeseries (which accounts for magnitude differences)
with the complexity of each timeseries (obtained through a measure of their variability) to
ensure reefs are clustered based on their underlying temporal dynamics rather than simply
their average coral cover. The process creates a distance matrix of n \* n dimensions (where
n is the number of timeseries in the group) that is clustered using a k-medoids clustering
algorithm [cf. @steinmann2020].

@@

To quantify the influence of reef characteristics on cluster assignment we used a Bayesian
ordinal mixed regression approach. Using Turing.jl [@fjelde2025] we model the influence of
five predictors, median depth, log total connectivity strength, log source to sink ratio,
mean DHW, log reef area and mean DHW tolerance level on cluster assignment, including random
intercepts for levels of **bioregion** and GCM identity.

@@

As one might expect, the number of years experience cover above $\beta_{0}$ (the assumed
positive carbonate budget threshold) substantially decreases as $\beta_{0}$ increases. The
distribution of number of years above carbonate-cover thresholds is right-skewed, with many
reefs experiencing low numbers of years above thresholds.

Reefs with greater median depth levels maintain coral cover above indicative carbonate
budget levels for a longer time frame compared to reefs with shallower median depths
[@fig-access-esm1-5-depth-carbonate]. Reef median depth has a moderate positive correlation
with the number of years a reef maintains coral cover above the required threshold
[@fig-access-esm1-5-depth-carbonate]. Depth - carbonate budget maintainence correlation
decreases as $\beta_{0}$ increases. This is likely due to an increasing number of reefs
being unable to maintain coral cover above threshold levels as the threshold increases, with
values becoming increasingly right-skewed. The base-10 log of total connectivity strength
was found to have no correlation with the number of years a reef maintains cover above the
required thresholds [@sfig-access-esm1-5-conn-carbonate]. Log connectivity strength -
carbonate budget maintainence correlation slightly decreases as the required carbonate
budget threshold increases.

>>>>

Using median coral cover timeseries for each reef we calculated the number of years each
reef maintains coral cover above a given percent threshold. The threshold represents the
level of live coral cover necessary to maintain a positive carbonate budget (denoted here as
$\theta_{0}$ to indicate the budget equilibrium). Greater number of years above $\theta_{0}$
therefore suggests greater resilience and opportunity for recovery.

Recent analyses suggest a general threshold of \~17% live coral cover in the case of the GBR
[@desbiens2021]. Given in situ carbonate budgets are influenced by (1) a range of processes;
(2) the variety of approaches that could be used to measure those processes; and (3) the
uncertainty surrounding these, it would be unreasonable to assume all reefs adhere to a
single precise $\theta_{0}$ value. Concurrently, it is difficult to ascertain a suitable
$\theta_{0}$ value for each individual reef.

Given the uncertainties present, the proportion of modelled reefs that display capacity to
maintain coral cover above a given $\theta_{0}$ threshold is examined. An indicative range
of 10 to 20% is used (inclusive, with a step size of 1%). The range was informed by Perry et
al., [-@perry2015] and Desbians et al., [-@desbiens2021] providing plausible indications of
the lower and upper bounds respectively. Robustness of the conclusions drawn to the
uncertainty in the precise value of $\theta_{0}$ for all reefs can then be assessed by using
the indicative range.

The relationship between the number of years above $\theta_{0}$ and median reef depth was
then assessed with Spearman's rank correlation. Spearman's correlation coefficient was
adopted as reef depth and connectivity strength distributions are expected to be non-normal
and highly skewed. Furthermore, reefs are naturally of different types and experience
different conditions, and so the monotonic trends (whether the relationship consistently
increase or decrease) rather than the exact quantative relationship was of interest for this
analysis.

@@

To identify groups of reefs that are resilient, intermediate, and non-resilient, we
clustered reefs into three clusters based on their temporal behaviour as indicated by coral
cover over time, using timeseries from 2025-2075. The Complexity-Invariant Distance (CID)
between each timeseries was calculated to identify timeseries that behave similarly
[@batista2014]. The CID corrects the Euclidean distance between pairs of timeseries (which
accounts for magnitude differences) with the complexity of each timeseries (obtained through
a measure of their variability) to ensure reefs are clustered based on their underlying
temporal dynamics rather than simply their average coral cover. The process creates a
distance matrix of n \* n dimensions (where n is the number of timeseries in the group) that
is clustered using a k-medoids clustering algorithm [cf. @steinmann2020]. The three reef
clusters are assigned a qualitative level from "low", "medium" and "high" based on their
relative median timeseries over the 2030-2050 period. Assigning cluster labels based on this
period prevents cluster labels being assigned during the steeply declining period from
2025-2030 and during the post-2050 period when many reefs have cover below 5% of their area.

@@

We used a Bayesian ordinal hierarchical regression model to quantify the influence of reef
characteristics on reef cluster assignment. Using Turing.jl [@fjelde2025] we model the
influence of five predictors, median depth, log total connectivity strength, log source to
sink ratio, mean DHW, log reef area and mean DHW tolerance level on cluster assignment,
including random intercepts for levels of **bioregion** and GCM identity. Based on initial
analyses suggesting higher cover clusters have reefs with greater depths, we constrain the
$\beta_{depth}$ coefficient prior to positive values. Similarly, we constrain the
$\beta_{DHW}$ coefficient prior to negative values as we do not expect increasing DHW levels to
increase cluster assignment level.

@@

As one might expect, the number of years experience cover above $\theta_{0}$ (the assumed
positive carbonate budget threshold) substantially decreases as $\theta_{0}$ increases. The
distribution of number of years above carbonate-cover thresholds is right-skewed, with many
reefs experiencing low numbers of years above thresholds.

Reefs with greater median depth levels maintain coral cover above indicative carbonate
budget levels for a longer time frame compared to reefs with shallower median depths
[@fig-access-esm1-5-depth-carbonate]. Reef median depth has a moderate positive correlation
with the number of years a reef maintains coral cover above the required threshold
[@fig-access-esm1-5-depth-carbonate]. Depth - carbonate budget maintainence correlation
decreases as $\theta_{0}$ increases. This is likely due to an increasing number of reefs
being unable to maintain coral cover above threshold levels as the threshold increases, with
values becoming increasingly right-skewed. The base-10 log of total connectivity strength
was found to have no correlation with the number of years a reef maintains cover above the
required thresholds [@sfig-access-esm1-5-conn-carbonate]. Log connectivity strength -
carbonate budget maintainence correlation slightly decreases as the required carbonate
budget threshold increases.
